---
title: "Natural Disaster Project Proposal"
author: "Sully, Adi, and Max"
output: html_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(ggplot2)
```

## 1. Introduction

In our project, we will aim to assess the evolution of natural disaster intensity, frequency, and location of the 20th century. We will generate figures that outline how climate change has accelerated these events over the last century. Our dataset comes from a database called Kaggle that provides a handful of climate datasets that are free to download. It was collected by compiling a set of historical natural disaster records to create a comprehensive history of natural disasters during the 20th century. The observations in our data set are individual natural disasters while the variables are characteristics of each disaster including Year, Disaster.Group, Country, Deaths, etc. Our research question is: In what ways have natural disasters intensified over the last century?

## 2. Data

```{r importing_data}
natural_disasters <- read.csv(file = "../data/DCS_Project.csv")

natural_disasters <- read.csv(file = "../data/DCS_Project.csv")

glimpse(natural_disasters)
```



## 3. Data analysis plan

1.    In this project, we are looking specifically at how the severity of natural disasters has changed over the last century (1900-2000). Some of the variables that will be of interest to us are disaster subgroup, type, and subtype to categorize the disaster, while variables like region, year, and deaths will give us characteristics of the disaster. We plan to use mostly bar plots and scatterplots to visualize the dataset. Most of our variables are catagorical, so density plots with specific fill variables may be useful as well.  

2.    We don't anticipate that any extra data will be necessary as there are enough columns in our dataset to create engaging figures.

3.    Below are a handful of figures that will be helpful in our first assessment of the data

```{r summary_statistics}
summary(natural_disasters)
str(natural_disasters)
```

Summarizing the data allows us to visualize the dataset quickly.

```{r density_visualizations}
ggplot(natural_disasters, mapping = aes(x = Year, fill = Disaster.Subgroup, alpha = 0.5)) +
  geom_density()
```
```{r visualizations_for_plot_critique}
filtered_data <- natural_disasters %>%
  filter(Year >= 1950 & Year <= 2025)

ggplot(filtered_data, mapping = aes(x = Year, fill = Disaster.Subgroup, alpha = 0.2, )) +
  geom_density()
```
```{r visualizations_for_plot_critique_2}

disaster_counts <- natural_disasters %>%
  count(Year, Disaster.Subgroup) %>%
  filter(Disaster.Subgroup != "Extra-terrestrial", Year >= 1940 & Year <= 2025)

pdf("Natural Disaster Heatmap.pdf")
ggplot(disaster_counts, aes(x = Year, y = Disaster.Subgroup, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "Heatmap of Disasters by Subgroup and Year", x = "Year", y = "Disaster Subgroup") +
  theme_minimal()

dev.off()
```
```{r visualizations_for_plot_critique_3}
pdf("my_plot.pdf", width = 8, height = 6)
ggplot(natural_disasters, aes(x = Year)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  facet_wrap(~ Disaster.Subgroup) +
  labs(title = "Disaster Years by Subgroup", x = "Year", y = "Count") +
  theme_minimal()
pdf()
```

```{r visualizations_for_plot_critique_4}
ggplot(natural_disasters, aes(x = Disaster.Subgroup)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Count of Disasters per Subgroup", x = "Disaster Subgroup", y = "Count") +
  theme_minimal()
```
```{r visualizations_for_plot_critique_5}
death_by_subgroup <- natural_disasters %>%
  group_by(Disaster.Subgroup) %>%
  summarise(Total.Deaths = sum(Total.Deaths, na.rm = TRUE))

# Create bar plot
ggplot(death_by_subgroup, aes(x = reorder(Disaster.Subgroup, Total.Deaths), y = Total.Deaths)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(title = "Total Deaths by Disaster Subgroup", x = "Disaster Subgroup", y = "Total Deaths") +
  coord_flip() + # Flips the axis for better readability
  theme_minimal()
```


Density plots can help us examine our dataset over the last century, as this is all data that has been collected since 1900. The plot above shows how the frequency of natural disasters has skyrocketed since the middle of the 20th century. 

```{r initial_visualizations}
ggplot(natural_disasters, mapping = aes(x = Total.Deaths, y = Total.Damages...000.US.., colour = Continent)) +
  geom_point() +
   xlim(min_value = 0, max_value = 500)+
   ylim(min_value = 0, max_value = 2.5e+07)
```

In the scatterplot above, we've plotted total deaths against total damages. This gives us a sense of what sort of relationship there is between fatalities and damages in these disasters. Additionally, we've added the continent variable as the color of the point on the scatterplot. This shows us that most of the disasters with the highest fatalities are in Africa or Asia while most of the American naturals disasters stay below 50 fatalities. 

## 4. Data Ethics Review

Limitations in Data Sources:

Our data comes from an open source website called Kaggle that provides datasets for download. Since our dataset is advertised as being a comprehensive log of all natural disasters from 1900-2021, there are plenty of opportunities for our data to exclude disasters or inaccurately record certain metrics. Additionally, much of the data was collected in a time when the technology we use to monitor these disasters wasn't quite what it is these days. This means that it is important to consider the difference in cost of damages, the possibility that some disasters were missed, or the chance that undeveloped areas of the world may have fatalities underestimated. 

Positive Effects on People:

It's possible that our project will be able to help us identify locations where people are most at risk of natural disasters in the future. Additionally, some of the world's most marginalized communities are disproportionately impacted by climate change. Our study could illuminate areas that are in greatest need of support.

Negative Effects on People:

The same way that identify at risk locations might help illuminate the areas of the world that are at risk moving forward, underestimating the severity or frequency of disasters in certain areas could categorize areas as safe when they are not. Again, undeveloped areas of the world likely see their natural disasters underestimated, while areas of the world that have the resources to monitor and clean up after a disaster will have their damages recorded accurately. 

Minimising Negative Impact:

Since this dataset might have omissions or misrepresentations, it is our responsibility to be on the lookout for patterns or errors that may have ended up in our figures. It would be useful to cross reference some of our data with records from other sources. In regards to damage costs, it would make sense to convert some of the early estimates to the 2025 USD scale, as the economy worked on a much smaller scale 100 years ago. 

